---
title: "Figure_2_stability_groups"
output: html_document
date: '2025-04-13'
editor_options: 
  chunk_output_type: console
---
#This script will regenerate Fig. 2C, E, G and Supplementary Fig. 2B
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = here::here("R", "Figure_2"))

merged_data_path <- here::here("data", "merged_data_GC2.tsv")
if (!file.exists(merged_data_path)) {
  stop("Missing input file: ", merged_data_path, call. = FALSE)
}
```

```{r}
library(dplyr);library(pheatmap);library(RColorBrewer);library(ggplot2);library(tidyr);library(patchwork)
```
# Load data
```{r}
merged_data_GC <- read.delim(merged_data_path, stringsAsFactors = FALSE)

merged_data_GC <- merged_data_GC %>% 
  mutate(
    Nuc_UT_mean = as.numeric(as.character(Nuc_UT_mean)),
    Nuc_30_mean = as.numeric(as.character(Nuc_30_mean)),
    Nuc_2h_mean = as.numeric(as.character(Nuc_2h_mean)),
    Nuc_4h_mean = as.numeric(as.character(Nuc_4h_mean)),
    Cyt_UT_mean = as.numeric(as.character(Cyt_UT_mean)),
    Cyt_30_mean = as.numeric(as.character(Cyt_30_mean)),
    Cyt_2h_mean = as.numeric(as.character(Cyt_2h_mean)),
    Cyt_4h_mean = as.numeric(as.character(Cyt_4h_mean)))

merged_data_GC <- merged_data_GC %>%
  mutate(
    Nuc_A_UT_IR_stability = Nuc_A_UT_1.x / Nuc_UT_mean,
    Nuc_B_UT_IR_stability = Nuc_B_UT_1.x / Nuc_UT_mean,
    
    Nuc_A_30_IR_stability = Nuc_A_30_1.x / Nuc_UT_mean,
    Nuc_B_30_IR_stability = Nuc_B_30_1.x / Nuc_UT_mean,
    
    Nuc_A_2h_IR_stability = Nuc_A_2h_1.x / Nuc_UT_mean,
    Nuc_B_2h_IR_stability = Nuc_B_2h_1.x / Nuc_UT_mean,
    
    Nuc_A_4h_IR_stability = Nuc_A_4h_1.x / Nuc_UT_mean,
    Nuc_B_4h_IR_stability = Nuc_B_4h_1.x / Nuc_UT_mean
  )
```

```{r}
# Create new columns comparing the stability of RNA in the nucleus
merged_data_GC_filtered <- merged_data_GC %>%
  mutate(Nuc_UT_TPM_stability = Nuc_UT_TPM_mean / Nuc_UT_TPM_mean,  # Reference of 1 for UT
         Nuc_30_TPM_stability = Nuc_30_TPM_mean / Nuc_UT_TPM_mean,
         Nuc_2h_TPM_stability = Nuc_2h_TPM_mean / Nuc_UT_TPM_mean,
         Nuc_4h_TPM_stability = Nuc_4h_TPM_mean / Nuc_UT_TPM_mean)

# Create new columns comparing the stability of RNA in the cytoplasm
merged_data_GC_filtered <- merged_data_GC_filtered %>%
  mutate(Cyt_UT_mean = as.numeric(as.character(Cyt_UT_mean)),
         Cyt_30_mean = as.numeric(as.character(Cyt_30_mean)),
         Cyt_2h_mean = as.numeric(as.character(Cyt_2h_mean)),
         Cyt_4h_mean = as.numeric(as.character(Cyt_4h_mean)))

merged_data_GC_filtered <- merged_data_GC_filtered %>%
  mutate(Cyt_UT_TPM_stability = Cyt_UT_TPM_mean / Cyt_UT_TPM_mean,  # Reference of 1 for UT
         Cyt_30_TPM_stability = Cyt_30_TPM_mean / Cyt_UT_TPM_mean,
         Cyt_2h_TPM_stability = Cyt_2h_TPM_mean / Cyt_UT_TPM_mean,
         Cyt_4h_TPM_stability = Cyt_4h_TPM_mean / Cyt_UT_TPM_mean)

# Create new columns comparing the stability of IR in the nucleus
merged_data_GC_filtered <- merged_data_GC_filtered %>%
  mutate(Nuc_UT_IR_stability = Nuc_UT_mean / Nuc_UT_mean,  # Reference of 1 for UT
         Nuc_30_IR_stability = Nuc_30_mean / Nuc_UT_mean,
         Nuc_2h_IR_stability = Nuc_2h_mean / Nuc_UT_mean,
         Nuc_4h_IR_stability = Nuc_4h_mean / Nuc_UT_mean)

# Create new columns comparing the stability of IR in the cytoplasm
merged_data_GC_filtered <- merged_data_GC_filtered %>%
  mutate(Cyt_UT_IR_stability = Cyt_UT_mean / Cyt_UT_mean,  # Reference of 1 for UT
         Cyt_30_IR_stability = Cyt_30_mean / Cyt_UT_mean,
         Cyt_2h_IR_stability = Cyt_2h_mean / Cyt_UT_mean,
         Cyt_4h_IR_stability = Cyt_4h_mean / Cyt_UT_mean)
```

#Filter out rows with 3 or more non-NAs in Nuc IR fractions
```{r}
merged_data_GC_filtered <- merged_data_GC_filtered %>%
  mutate(non_na_count = rowSums(!is.na(dplyr::select(., Nuc_A_UT_1.x, Nuc_B_UT_1.x, Nuc_A_30_1.x, Nuc_B_30_1.x, Nuc_A_2h_1.x, Nuc_B_2h_1.x, Nuc_A_4h_1.x, Nuc_B_4h_1.x)))) %>%
  filter(non_na_count >= 5)

```
# For gene_orientated studies sometimes you need to filter the data to keep only the intron with the highest Nuc_UT_mean for each gene
```{r}
merged_data_GC_filtered_GENE <- merged_data_GC_filtered %>%
  group_by(GENE) %>%
  top_n(1, Nuc_UT_mean) %>%  # Select the top 1 row with the highest Nuc_UT_mean per GENE
  ungroup()
```

# Assign the stability to introns
```{r}
#--------------------------------------------#
#### Assign stability, here I care about introns so I dont need to filter overlaping genes
#--------------------------------------------#
IR_UT <- merged_data_GC_filtered %>%
  filter(
    Nuc_UT_mean >= 30 & 
      (
        (Nuc_A_UT_1.x >= 25 & Nuc_B_UT_1.x >= 25) |
        (is.na(Nuc_A_UT_1.x) & Nuc_B_UT_1.x >= 25) |
        (Nuc_A_UT_1.x >= 25 & is.na(Nuc_B_UT_1.x))
      )
  ) %>%
  mutate(stability = "IR")

stableR4h <- merged_data_GC_filtered %>%
  filter(
    Nuc_UT_mean >= 30 &
    Nuc_4h_IR_stability >= 0.5 &
    Nuc_30_IR_stability >= 0.5 &
    Nuc_2h_IR_stability >= 0.5 &
    (
      (Nuc_A_UT_1.x >= 25 & Nuc_B_UT_1.x >= 25) |
      (is.na(Nuc_A_UT_1.x) & Nuc_B_UT_1.x >= 25) |
      (Nuc_A_UT_1.x >= 25 & is.na(Nuc_B_UT_1.x))
    ) &
    Nuc_30_mean >= 25 &
    Nuc_2h_mean >= 25 &
    Nuc_4h_mean >= 25
  ) %>% mutate(stability = "stable 4h")


unstableR30min <- merged_data_GC_filtered %>%
  filter(
    Nuc_UT_mean >= 30 &
    Nuc_30_IR_stability < 0.5 &
    Nuc_2h_IR_stability < 0.5 &
    Nuc_4h_IR_stability < 0.5 &
    (
      (Nuc_A_UT_1.x >= 25 & Nuc_B_UT_1.x >= 25) |
      (is.na(Nuc_A_UT_1.x) & Nuc_B_UT_1.x >= 25) |
      (Nuc_A_UT_1.x >= 25 & is.na(Nuc_B_UT_1.x))
    )) %>%
  mutate(stability = "unstable30min")


unstableR2h <- merged_data_GC_filtered %>%
  filter(
    Nuc_UT_mean >= 30 &
    Nuc_30_IR_stability >= 0.5 &
    Nuc_2h_IR_stability < 0.5 &
    Nuc_4h_IR_stability < 0.5 &
    Nuc_30_mean >= 25 &
    (
      (Nuc_A_UT_1.x >= 25 & Nuc_B_UT_1.x >= 25) |
      (is.na(Nuc_A_UT_1.x) & Nuc_B_UT_1.x >= 25) |
      (Nuc_A_UT_1.x >= 25 & is.na(Nuc_B_UT_1.x)))
  ) %>% mutate(stability = "unstable2h")

unstableR4h <- merged_data_GC_filtered %>%
  filter(
    Nuc_UT_mean >= 30 &
    Nuc_4h_IR_stability < 0.5 &
    Nuc_30_IR_stability >= 0.5 &
    Nuc_2h_IR_stability >= 0.5 &
    Nuc_30_mean >= 25 &  
    Nuc_2h_mean >= 25 & 
    (
      (Nuc_A_UT_1.x >= 25 & Nuc_B_UT_1.x >= 25) |
      (is.na(Nuc_A_UT_1.x) & Nuc_B_UT_1.x >= 25) |
      (Nuc_A_UT_1.x >= 25 & is.na(Nuc_B_UT_1.x))
    )
  ) %>% mutate(stability = "unstable 4h")

noIR <- merged_data_GC_filtered %>%
  filter(Nuc_UT_mean < 30) %>%
  mutate(stability = "no IR")

##no IR GENE
noIR_GENE <- merged_data_GC_filtered_GENE %>%
  filter((Nuc_UT_mean) < 30) %>%
  mutate(stability = "no_IR_GENE")

noIR_GENE <- noIR_GENE %>%
  distinct(GENE, .keep_all = TRUE)
sum(is.na(noIR_GENE$Nuc_UT_mean))

combined_data <- bind_rows(stableR4h, unstableR4h, unstableR30min, unstableR2h, noIR)
```

###same settings at 50%
```{r}
# IR at UT (50 cutoff)
IR_UT_50cut <- subset(merged_data_GC_filtered,
                      Nuc_UT_mean >= 50 &
                      (
                        (Nuc_A_UT_1.x >= 50 & Nuc_B_UT_1.x >= 50) |
                        (is.na(Nuc_A_UT_1.x) & Nuc_B_UT_1.x >= 50) |
                        (Nuc_A_UT_1.x >= 50 & is.na(Nuc_B_UT_1.x))
                      )) %>% mutate(stability = "IR")

# Stable at 4h (50 cutoff)
stableR4h_50cut <- subset(
  merged_data_GC_filtered,
  Nuc_UT_mean >= 50 &
  Nuc_4h_IR_stability >= 0.5 &
  Nuc_30_IR_stability >= 0.5 &
  Nuc_2h_IR_stability >= 0.5 
  ) %>%
  mutate(stability = "stable 4h")

# Unstable at 30 min (50 cutoff)
unstableR30min_50cut <- subset(
  merged_data_GC_filtered,
  Nuc_UT_mean >= 50 &
  Nuc_30_IR_stability < 0.5 &
  Nuc_2h_IR_stability < 0.5 &
  Nuc_4h_IR_stability < 0.5
  ) %>%
  mutate(stability = "unstable30min")

# Unstable at 2h (50 cutoff)
unstableR2h_50cut <- subset(
  merged_data_GC_filtered,
  Nuc_UT_mean >= 50 &
  Nuc_30_IR_stability >= 0.5 &
  Nuc_2h_IR_stability < 0.5 &
  Nuc_4h_IR_stability < 0.5 #&
  ) %>%
  mutate(stability = "unstable2h") 

# Unstable at 4h (50 cutoff)
unstableR4h_50cut <- subset(
  merged_data_GC_filtered,
  Nuc_UT_mean >= 50 &
  Nuc_4h_IR_stability < 0.5 &
  Nuc_30_IR_stability >= 0.5 &
  Nuc_2h_IR_stability >= 0.5 #&
) %>% mutate(stability = "unstable 4h") 

noIR_50cut <- merged_data_GC_filtered %>%
  filter(Nuc_UT_mean < 50) %>%
  mutate(stability = "no IR")

# Final combined dataset for 50% IR cutoff
combined_data_50cut <- bind_rows(
  stableR4h_50cut,
  unstableR4h_50cut,
  unstableR30min_50cut,
  unstableR2h_50cut,
  noIR_50cut
)
```

### Intron differences with different cutoff criteria 30%:
```{r}
combined_dataTPMno100 <- combined_data %>%
  filter(Nuc_UT_TPM_mean >= 5) %>% 
  filter(is.na(Cyt_A_UT_1.x) | is.na(Cyt_B_UT_1.x) | #keep is.na because otherwise they get moved
  (Cyt_A_UT_1.x < 99 & Cyt_B_UT_1.x < 99))

summary_table_no100 <- combined_dataTPMno100 %>%
  group_by(stability) %>%  # Group the data by the stability group
  summarise(
    count         = n(),  # Number of observations per group
    mean_GC       = mean(GC_Content, na.rm = TRUE),
    median_GC     = median(GC_Content, na.rm = TRUE),
    sd_GC         = sd(GC_Content, na.rm = TRUE),
    min_GC        = min(GC_Content, na.rm = TRUE),
    max_GC        = max(GC_Content, na.rm = TRUE),
    mean_LENGTH   = mean(LENGTH, na.rm = TRUE),
    median_LENGTH = median(LENGTH, na.rm = TRUE),
    sd_LENGTH     = sd(LENGTH, na.rm = TRUE),
    min_LENGTH    = min(LENGTH, na.rm = TRUE),
    max_LENGTH    = max(LENGTH, na.rm = TRUE))

## Check filtering didn't affect the results
print(summary_table_no100)

# Filter the two groups of interest
group1 <- combined_dataTPMno100 %>% filter(stability == "stable 4h")
group2 <- combined_dataTPMno100 %>% filter(stability == "unstable30min")

# Define base data
base_group <- combined_dataTPMno100 %>% filter(stability == "unstable30min")
# Initialize list to store results
length_tests <- list()

# Define comparison groups
comparison_groups <- c("stable 4h", "unstable2h", "unstable 4h")

# Loop through each group and perform Wilcoxon test vs unstable30min
for (grp in comparison_groups) {
  group <- combined_dataTPMno100 %>% filter(stability == grp)

  test_result <- wilcox.test(group$LENGTH, base_group$LENGTH, alternative = "two.sided", exact = FALSE)

  length_tests[[grp]] <- list(
    group_vs_unstable30min = grp,
    p_value = test_result$p.value,
    statistic = test_result$statistic,
    method = test_result$method
  )
}

# Convert results to data frame for display
length_results_df <- do.call(rbind, lapply(length_tests, as.data.frame))
print(length_results_df)

# Wilcoxon test for GC content
gc_test <- wilcox.test(group1$GC_Content, group2$GC_Content, alternative = "two.sided", exact = FALSE)

# Wilcoxon test for intron length
length_test <- wilcox.test(group1$LENGTH, group2$LENGTH, alternative = "two.sided", exact = FALSE)

# Print results
cat("GC Content - Wilcoxon Test:\n")
print(gc_test)

cat("\nIntron Length - Wilcoxon Test:\n")
print(length_test)

# Define base data
base_group <- combined_dataTPMno100 %>% filter(stability == "unstable30min")

# Initialize list to store results
gc_tests <- list()

# Define comparison groups
comparison_groups <- c("stable 4h", "unstable2h", "unstable 4h")

# Loop through each group and perform Wilcoxon test vs unstable30min
for (grp in comparison_groups) {
  group <- combined_dataTPMno100 %>% filter(stability == grp)

  test_result <- wilcox.test(group$GC_Content, base_group$GC_Content, alternative = "two.sided", exact = FALSE)

  gc_tests[[grp]] <- list(
    group_vs_unstable30min = grp,
    p_value = test_result$p.value,
    statistic = test_result$statistic,
    method = test_result$method
  )
}

# Convert results to data frame for display
gc_results_df <- do.call(rbind, lapply(gc_tests, as.data.frame))
print(gc_results_df)
```

###same summary tables for 50% 
```{r}
# Create TPM-filtered datasets from 50% IR cutoff classifications
combined_dataTPMno100_50cut <- combined_data_50cut %>%
  filter(Nuc_UT_TPM_mean >= 5) %>%
   filter(is.na(Cyt_A_UT_1.x) | is.na(Cyt_B_UT_1.x) | 
         (Cyt_A_UT_1.x < 99 & Cyt_B_UT_1.x < 99))

summary_table_no100_50cut <- combined_dataTPMno100_50cut %>%
  group_by(stability) %>%
  summarise(
    count         = n(),
    mean_GC       = mean(GC_Content, na.rm = TRUE),
    median_GC     = median(GC_Content, na.rm = TRUE),
    sd_GC         = sd(GC_Content, na.rm = TRUE),
    min_GC        = min(GC_Content, na.rm = TRUE),
    max_GC        = max(GC_Content, na.rm = TRUE),
    mean_LENGTH   = mean(LENGTH, na.rm = TRUE),
    median_LENGTH = median(LENGTH, na.rm = TRUE),
    sd_LENGTH     = sd(LENGTH, na.rm = TRUE),
    min_LENGTH    = min(LENGTH, na.rm = TRUE),
    max_LENGTH    = max(LENGTH, na.rm = TRUE)
  )

# Print summaries, check filtering didnt affect the results
print(summary_table_no100_50cut)

# Wilcoxon test between two IR stability groups (50% cutoff)
group1_50cut <- combined_dataTPMno100_50cut %>% filter(stability == "stable 4h")
group2_50cut <- combined_dataTPMno100_50cut %>% filter(stability == "unstable30min")

# GC content test
gc_test_50cut <- wilcox.test(group1_50cut$GC_Content, group2_50cut$GC_Content,
                             alternative = "two.sided", exact = FALSE)

# Intron length test
length_test_50cut <- wilcox.test(group1_50cut$LENGTH, group2_50cut$LENGTH,
                                 alternative = "two.sided", exact = FALSE)

# Print test results
cat("GC Content - Wilcoxon Test (50% cutoff):\n")
print(gc_test_50cut)

cat("\nIntron Length - Wilcoxon Test (50% cutoff):\n")
print(length_test_50cut)

# Initialize list to store results
gc_tests <- list()

# Define comparison groups
comparison_groups <- c("stable 4h", "unstable2h", "unstable 4h")

# Define the base group for comparison
base_group <- combined_dataTPMno100_50cut %>%
  filter(stability == "unstable30min")

# Loop through each group and perform Wilcoxon test vs unstable30min
for (grp in comparison_groups) {
  group <- combined_dataTPMno100_50cut %>%
    filter(stability == grp)
  
  test_result <- wilcox.test(group$GC_Content, base_group$GC_Content,
                             alternative = "two.sided", exact = FALSE)
  
  gc_tests[[grp]] <- list(
    group_vs_unstable30min = grp,
    p_value = test_result$p.value,
    statistic = test_result$statistic,
    method = test_result$method
  )
}

# Convert results to a data frame for display
gc_results_df_50cut <- do.call(rbind, lapply(gc_tests, as.data.frame))

# Print the results
print(gc_results_df_50cut)
```

### plot for IR stability over time, length, PIR UT, GC percentage at PIR 30% cutoff
```{r}
# Define custom colors for timepoints
myColors <- colorRampPalette(c("darkred", "darkgray"))(4)
myStabColors <- colorRampPalette(c("darkred", "darkgray"))(4)

# Prepare long-format data for violin plots
combined_violin_long <- combined_dataTPMno100 %>%
  filter(stability %in% c("unstable30min", "unstable2h", "unstable 4h", "stable 4h")) %>%
  dplyr::select(GENE, stability,
                Nuc_UT_mean, Nuc_30_mean, Nuc_2h_mean, Nuc_4h_mean) %>%
  pivot_longer(
    cols = starts_with("Nuc_"),
    names_to = "TimePoint",
    values_to = "NuclearMean") %>%
  mutate(
    TimePoint = recode(TimePoint,
                       "Nuc_UT_mean" = "UT",
                       "Nuc_30_mean" = "30min",
                       "Nuc_2h_mean" = "2h",
                       "Nuc_4h_mean" = "4h"),
    TimePoint = factor(TimePoint, levels = c("UT", "30min", "2h", "4h")),
    stability = factor(stability, levels = c("unstable30min", "unstable2h", "unstable 4h", "stable 4h")))

# Calculate a shared y-axis range
ymax <- max(combined_violin_long$NuclearMean, na.rm = TRUE)
ybreaks <- pretty(c(0, ymax))
upper_limit <- max(ybreaks) + 10  # Add extra space above the top tick

# Calculate means per TimePoint and stability group
mean_points <- combined_violin_long %>%
  group_by(stability, TimePoint) %>%
  summarise(mean_val = mean(NuclearMean, na.rm = TRUE)) %>%
  ungroup()

# Generate the violin plot with shared scale and extra headroom
p <- ggplot(combined_violin_long, aes(x = TimePoint, y = NuclearMean, fill = TimePoint)) +
  geom_violin(trim = FALSE, scale = "width", alpha = 0.7) + #if you want the width to represent actual data density, then geom_violin(trim = FALSE, alpha = 0.7)
  geom_boxplot(width = 0.1, outlier.shape = NA, fill = "white") +
  facet_wrap(~ stability, ncol = 2, scales = "fixed") +
  scale_y_continuous(breaks = ybreaks, limits = c(min(ybreaks), upper_limit)) +
  scale_fill_manual(values = myColors) +
  labs(title = "Violin Plots of Nuclear IR Means by Time Point",
       x = "Time Point", y = "Nuclear Mean PIR") +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    strip.text = element_text(face = "bold", size = 13)
  )

p

# Calculate mean NuclearMean PIR per time point and stability group
pir_summary <- combined_violin_long %>%
  group_by(stability, TimePoint) %>%
  summarise(
    mean_PIR = mean(NuclearMean, na.rm = TRUE),
    median_PIR = median(NuclearMean, na.rm = TRUE),
    sd_PIR = sd(NuclearMean, na.rm = TRUE),
    n = n()
  ) %>%
  arrange(stability, TimePoint)

# View the table
print(pir_summary)

###NUCLEAR IR levels

# Calculate y-axis breaks and upper limit
ymax <- max(combined_dataTPMno100$Nuc_UT_mean, na.rm = TRUE)
ybreaks <- pretty(c(0, ymax))
upper_limit <- max(ybreaks) + 10

p_nuc_ut <- combined_dataTPMno100 %>%
  filter(stability %in% c("unstable30min", "unstable2h", "unstable 4h", "stable 4h")) %>%
  mutate(stability = factor(stability, levels = c("unstable30min", "unstable2h", "unstable 4h", "stable 4h"))) %>%
  ggplot(aes(x = stability, y = Nuc_UT_mean, fill = stability)) +
  geom_violin(trim = FALSE, alpha = 0.8) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  labs(title = "Nuclear IR (UT)", x = "Stability Group", y = "Nuc_UT_mean") +
  scale_y_continuous(breaks = ybreaks, limits = c(0, upper_limit)) +
  scale_fill_manual(values = myStabColors) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    axis.text = element_text(color = "black"),
    legend.position = "none"
  )

p_nuc_ut

## LENGTH
p_length_log <- combined_dataTPMno100 %>%
  filter(stability %in% c("unstable30min", "unstable2h", "unstable 4h", "stable 4h"),
         LENGTH > 0) %>%  # Make sure there are no 0-length values
  mutate(
    stability = factor(stability, levels = c("unstable30min", "unstable2h", "unstable 4h", "stable 4h"))
  ) %>%
  ggplot(aes(x = stability, y = LENGTH, fill = stability)) +
  geom_violin(trim = FALSE, alpha = 0.8) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  scale_y_log10(limits = c(10, NA), breaks = c(10, 100, 1000, 10000, 100000)) +
  labs(title = "Intron Length (log10 scale)", x = "Stability Group", y = "Length (log10 nt)") +
  scale_fill_manual(values = myStabColors) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

p_length_log

####GC %
ymax_gc <- max(combined_dataTPMno100$GC_Content, na.rm = TRUE)
ybreaks_gc <- pretty(c(0, ymax_gc))
upper_limit_gc <- max(ybreaks_gc, 100)  # Ensure it reaches at least 100

ymax <- max(combined_dataTPMno100$Nuc_UT_mean, na.rm = TRUE)
ybreaks <- pretty(c(0, ymax))
upper_limit <- max(ybreaks) + 10
# Plot
p_gc <- combined_dataTPMno100 %>%
  filter(stability %in% c("unstable30min", "unstable2h", "unstable 4h", "stable 4h")) %>%
  mutate(stability = factor(stability, levels = c("unstable30min", "unstable2h", "unstable 4h", "stable 4h"))) %>%
  ggplot(aes(x = stability, y = GC_Content, fill = stability)) +
  geom_violin(trim = FALSE, alpha = 0.8) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  scale_y_continuous(breaks = ybreaks_gc, limits = c(0, upper_limit)) +
  labs(title = "GC Content", x = "Stability Group", y = "GC (%)") +
  scale_fill_manual(values = myStabColors) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

p_gc

combined_plot <- (p_nuc_ut | p_gc) /
                 (p_length_log | p3) #p3 is generated after selecting GENES later
combined_plot
```

#the same plot for 50% cutoff
```{r}
### 1. Prepare long-format data for violin plots
combined_violin_long <- combined_dataTPMno100_50cut %>%
  filter(stability %in% c("unstable30min", "unstable2h", "unstable 4h", "stable 4h")) %>%
  dplyr::select(GENE, stability, Nuc_UT_mean, Nuc_30_mean, Nuc_2h_mean, Nuc_4h_mean) %>%
  pivot_longer(
    cols = starts_with("Nuc_"),
    names_to = "TimePoint",
    values_to = "NuclearMean"
  ) %>%
  mutate(
    TimePoint = recode(TimePoint,
                       "Nuc_UT_mean" = "UT",
                       "Nuc_30_mean" = "30min",
                       "Nuc_2h_mean" = "2h",
                       "Nuc_4h_mean" = "4h"),
    TimePoint = factor(TimePoint, levels = c("UT", "30min", "2h", "4h")),
    stability = factor(stability, levels = c("unstable30min", "unstable2h", "unstable 4h", "stable 4h"))
  )

# Calculate shared y-axis
ymax <- max(combined_violin_long$NuclearMean, na.rm = TRUE)
ybreaks <- pretty(c(0, ymax))
upper_limit <- max(ybreaks) + 10

# Calculate means for dot overlay
mean_points <- combined_violin_long %>%
  group_by(stability, TimePoint) %>%
  summarise(mean_val = mean(NuclearMean, na.rm = TRUE)) %>%
  ungroup()

# Plot nuclear PIR violin plot
p <- ggplot(combined_violin_long, aes(x = TimePoint, y = NuclearMean, fill = TimePoint)) +
  geom_violin(trim = FALSE, scale = "width", alpha = 0.7) +
  geom_boxplot(width = 0.1, outlier.shape = NA, fill = "white") +
  facet_wrap(~ stability, ncol = 2, scales = "fixed") +
  scale_y_continuous(breaks = ybreaks, limits = c(min(ybreaks), upper_limit)) +
  scale_fill_manual(values = myColors) +
  labs(title = "Violin Plots of Nuclear IR Means by Time Point",
       x = "Time Point", y = "Nuclear Mean PIR") +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    strip.text = element_text(face = "bold", size = 13)
  )

# Show plot
p

# Summary table
pir_summary_50cut <- combined_violin_long %>%
  group_by(stability, TimePoint) %>%
  summarise(
    mean_PIR = mean(NuclearMean, na.rm = TRUE),
    median_PIR = median(NuclearMean, na.rm = TRUE),
    sd_PIR = sd(NuclearMean, na.rm = TRUE),
    n = n()
  ) %>%
  arrange(stability, TimePoint)

print(pir_summary_50cut)

### 2. Basal Nuclear IR Plot
ymax <- max(combined_dataTPMno100_50cut$Nuc_UT_mean, na.rm = TRUE)
ybreaks <- pretty(c(0, ymax))
upper_limit <- max(ybreaks) + 10

p_nuc_ut <- combined_dataTPMno100_50cut %>%
  filter(stability %in% c("unstable30min", "unstable2h", "unstable 4h", "stable 4h")) %>%
  mutate(stability = factor(stability, levels = c("unstable30min", "unstable2h", "unstable 4h", "stable 4h"))) %>%
  ggplot(aes(x = stability, y = Nuc_UT_mean, fill = stability)) +
  geom_violin(trim = FALSE, alpha = 0.8) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  labs(title = "Nuclear IR (UT)", x = "Stability Group", y = "Nuc_UT_mean") +
  scale_y_continuous(breaks = ybreaks, limits = c(0, upper_limit)) +
  scale_fill_manual(values = myStabColors) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    axis.text = element_text(color = "black"),
    legend.position = "none"
  )

### 3. Intron Length (log10)
p_length_log <- combined_dataTPMno100_50cut %>%
  filter(stability %in% c("unstable30min", "unstable2h", "unstable 4h", "stable 4h"),
         LENGTH > 0) %>%
  mutate(stability = factor(stability, levels = c("unstable30min", "unstable2h", "unstable 4h", "stable 4h"))) %>%
  ggplot(aes(x = stability, y = LENGTH, fill = stability)) +
  geom_violin(trim = FALSE, alpha = 0.8) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  scale_y_log10(limits = c(10, NA), breaks = c(10, 100, 1000, 10000, 100000)) +
  labs(title = "Intron Length (log10 scale)", x = "Stability Group", y = "Length (log10 nt)") +
  scale_fill_manual(values = myStabColors) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

### 4. GC Content
ymax_gc <- max(combined_dataTPMno100_50cut$GC_Content, na.rm = TRUE)
ybreaks_gc <- pretty(c(0, ymax_gc))
upper_limit_gc <- max(ybreaks_gc, 100)

p_gc <- combined_dataTPMno100_50cut %>%
  filter(stability %in% c("unstable30min", "unstable2h", "unstable 4h", "stable 4h")) %>%
  mutate(stability = factor(stability, levels = c("unstable30min", "unstable2h", "unstable 4h", "stable 4h"))) %>%
  ggplot(aes(x = stability, y = GC_Content, fill = stability)) +
  geom_violin(trim = FALSE, alpha = 0.8) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  scale_y_continuous(breaks = ybreaks_gc, limits = c(0, upper_limit_gc)) +
  labs(title = "GC Content", x = "Stability Group", y = "GC (%)") +
  scale_fill_manual(values = myStabColors) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

# Combine plots
combined_plot_50cut <- (p_nuc_ut | p_gc) /
                 (p_length_log | p3_50cut) # p3_50cut is generated after selecting GENES later
```

#############
#generating stable and unstable GENES, 30% cutoff
#############
```{r}
#Check overlapping introns between groups
sets_list <- list(
  "Stable 4h" = unique(stableR4h$COORD),
  "Unstable 4h" = unique(unstableR4h$COORD),
  "Unstable 30min" = unique(unstableR30min$COORD),
  "Unstable 2h" = unique(unstableR2h$COORD),
  "No IR" = unique(noIR$COORD))

### Check overlap between GENES in different groups ####
# First, make a list of genes for each stability category of introns
sets_list <- list(
  "Stable 4h" = unique(stableR4h$GENE),
  "Unstable 4h" = unique(unstableR4h$GENE),
  "Unstable 30min" = unique(unstableR30min$GENE),
  "Unstable 2h" = unique(unstableR2h$GENE))

###------------------------------------------------------------------------------####
# ---- Now eliminate overlapping GENES favoring the longer stability group -----######
# ---- Option 1: Extract exclusive genes from each set individually ----
exclusive_by_set <- lapply(names(sets_list), function(set_name) {
  current_set <- sets_list[[set_name]]
  # Combine all the genes in the other sets into one vector:
  other_sets <- unlist(sets_list[names(sets_list) != set_name])
  setdiff(current_set, other_sets)})
names(exclusive_by_set) <- names(sets_list)

# View the exclusive (non-overlapping) genes in each category:
exclusive_by_set  # This is a list where each element contains the non-overlapping genes for that category
exclusive_by_set[["Unstable 30min"]]

# If you want a union of all genes that appear in only one category, you can combine them:
non_overlapping_genes <- unlist(exclusive_by_set)
non_overlapping_genes

### CLEANING over lapping genes ###
exclusive_unstable30min <- setdiff(
  sets_list[["Unstable 30min"]],
  union(union(sets_list[["Stable 4h"]], sets_list[["Unstable 4h"]]), sets_list[["Unstable 2h"]]))

exclusive_unstable2h <- setdiff(
  sets_list[["Unstable 2h"]],
  union(sets_list[["Stable 4h"]], sets_list[["Unstable 4h"]]))

exclusive_unstable4h <- setdiff(unstableR4h$GENE, stableR4h$GENE)

### STEP 2. Check if GENES still overlapp
sets_list_updated <- list(
  "Stable 4h"      = unique(stableR4h$GENE),
  "Unstable 4h"    = unique(exclusive_unstable4h),
  "Unstable 30min" = unique(exclusive_unstable30min),
  "Unstable 2h"    = unique(exclusive_unstable2h))

### STEP 4. Compare the Union of Genes Before and After Filtering
initial_genes <- unique(c(
  stableR4h$GENE,
  unstableR4h$GENE,
  unstableR30min$GENE,
  unstableR2h$GENE))

# For the filtered union, we use the updated sets (note that unstable4h_exclusive is part of the union)
filtered_genes <- unique(c(
  stableR4h$GENE,
  exclusive_unstable4h,              # Already removed genes overlapping stableR4h
  exclusive_unstable30min,
  exclusive_unstable2h))

# Print out the total counts for comparison
cat("Total unique genes before filtering:", length(initial_genes), "\n")
cat("Total unique genes after filtering: ", length(filtered_genes), "\n")

# You can also check that the sets are identical (if the overall union remains unchanged)
if (setequal(initial_genes, filtered_genes)) {
  cat("The union of genes is identical before and after filtering.\n")
} else {
  cat("The union of genes differs between the initial and filtered sets.\n")}

# Create labeled data frame from stableR4h (which is a data frame)
df_stable4h <- stableR4h %>%
  dplyr::select(GENE) %>%
  mutate(GENE_stability = "stable_4h")

# For the others, turn them into data frames directly
df_unstable4h <- data.frame(GENE = exclusive_unstable4h,
                            GENE_stability = "unstable_4h")

df_unstable30min <- data.frame(GENE = exclusive_unstable30min,
                               GENE_stability = "unstable_30min")

df_unstable2h <- data.frame(GENE = exclusive_unstable2h,
                            GENE_stability = "unstable_2h")

# Combine all into one data frame
gene_stability_df <- bind_rows(
  df_stable4h,
  df_unstable4h,
  df_unstable30min,
  df_unstable2h)
gene_stability_df_unique <- gene_stability_df %>%
  distinct(GENE, .keep_all = TRUE)  # Keep only one row per gene

```

###for 50% cutoff stable and unstable GENEs
```{r}
# COORD-level overlap (50% cutoff)
sets_list_50cut_coord <- list(
  "Stable 4h"       = unique(stableR4h_50cut$COORD),
  "Unstable 4h"     = unique(unstableR4h_cleaned_50cut$COORD),
  "Unstable 30min"  = unique(unstableR30min_cleaned_50cut$COORD),
  "Unstable 2h"     = unique(unstableR2h_cleaned_50cut$COORD),
  "No IR"           = unique(noIR_50cut$COORD)
)

# GENE-level overlap (50% cutoff)
sets_list_50cut_gene <- list(
  "Stable 4h"       = unique(stableR4h_50cut$GENE),
  "Unstable 4h"     = unique(unstableR4h_50cut$GENE),
  "Unstable 30min"  = unique(unstableR30min_50cut$GENE),
  "Unstable 2h"     = unique(unstableR2h_50cut$GENE)
  # "No IR" can be included if desired
)

# Exclusive genes per category
exclusive_by_set_50cut <- lapply(names(sets_list_50cut_gene), function(set_name) {
  current_set <- sets_list_50cut_gene[[set_name]]
  other_sets <- unlist(sets_list_50cut_gene[names(sets_list_50cut_gene) != set_name])
  setdiff(current_set, other_sets)
})
names(exclusive_by_set_50cut) <- names(sets_list_50cut_gene)

# Direct access
exclusive_by_set_50cut[["Unstable 30min"]]

# Union of all non-overlapping genes
non_overlapping_genes_50cut <- unlist(exclusive_by_set_50cut)

# Manual cleaning (preferred longer stability)
exclusive_unstable30min_50cut <- setdiff(
  sets_list_50cut_gene[["Unstable 30min"]],
  union(union(sets_list_50cut_gene[["Stable 4h"]], sets_list_50cut_gene[["Unstable 4h"]]),
        sets_list_50cut_gene[["Unstable 2h"]])
)

exclusive_unstable2h_50cut <- setdiff(
  sets_list_50cut_gene[["Unstable 2h"]],
  union(sets_list_50cut_gene[["Stable 4h"]], sets_list_50cut_gene[["Unstable 4h"]])
)

exclusive_unstable4h_50cut <- setdiff(
  unstableR4h_50cut$GENE,
  stableR4h_50cut$GENE
)

# Updated gene sets after filtering
sets_list_updated_50cut <- list(
  "Stable 4h"      = unique(stableR4h_50cut$GENE),
  "Unstable 4h"    = unique(exclusive_unstable4h_50cut),
  "Unstable 30min" = unique(exclusive_unstable30min_50cut),
  "Unstable 2h"    = unique(exclusive_unstable2h_50cut)
)

# Compare gene set sizes before/after filtering
initial_genes_50cut <- unique(c(
  stableR4h_50cut$GENE,
  unstableR4h_50cut$GENE,
  unstableR30min_50cut$GENE,
  unstableR2h_50cut$GENE
))

filtered_genes_50cut <- unique(c(
  stableR4h_50cut$GENE,
  exclusive_unstable4h_50cut,
  exclusive_unstable30min_50cut,
  exclusive_unstable2h_50cut
))

cat("Total unique genes before filtering (50%):", length(initial_genes_50cut), "\n")
cat("Total unique genes after filtering (50%): ", length(filtered_genes_50cut), "\n")

if (setequal(initial_genes_50cut, filtered_genes_50cut)) {
  cat("The union of genes is identical before and after filtering.\n")
} else {
  cat("The union of genes differs between the initial and filtered sets.\n")
}

# Create labeled gene data frame
df_stable4h_50cut <- stableR4h_50cut %>%
  dplyr::select(GENE) %>%
  mutate(GENE_stability = "stable_4h")

df_unstable4h_50cut <- data.frame(GENE = exclusive_unstable4h_50cut,
                                  GENE_stability = "unstable_4h")

df_unstable30min_50cut <- data.frame(GENE = exclusive_unstable30min_50cut,
                                     GENE_stability = "unstable_30min")

df_unstable2h_50cut <- data.frame(GENE = exclusive_unstable2h_50cut,
                                  GENE_stability = "unstable_2h")

# Final gene stability annotation table
gene_stability_df_50cut <- bind_rows(
  df_stable4h_50cut,
  df_unstable4h_50cut,
  df_unstable30min_50cut,
  df_unstable2h_50cut
)

gene_stability_df_unique_50cut <- gene_stability_df_50cut %>%
  distinct(GENE, .keep_all = TRUE)

```

### LINE PLOT STABILITY, GO AND NUC ENRICHMENT 30% ###
```{r}
# --- STEP 1: Filter the Combined Data for Exclusive Genes ---
##Use these filtering conditions for RNA STABILITY in the nucleus and in the cytoplasm. In order to monitor RNA stability in the cytoplasm we filtered lowly expressed genes in the cytoplasm, RAD52 smRNA FISH was used to set the minimum threshold, which among our IR mRNAs exhibited the lowest cytoplasmic abundance (mean of one molecule per cytoplasm per cell), corresponded to 7 TPM in the cytoplasm. 
combined_dataTPMno100 <- combined_data %>%
 filter(Nuc_UT_TPM_mean >= 7 & Cyt_UT_TPM_mean >= 7) %>%
  filter(is.na(Cyt_A_UT_1.x) | is.na(Cyt_B_UT_1.x) | #keep is.na because otherwise they get moved
   (Cyt_A_UT_1.x < 99 & Cyt_B_UT_1.x < 99))

combined_exclusive <- combined_dataTPMno100 %>%
  filter(GENE %in% gene_stability_df_unique$GENE) %>%
  left_join(gene_stability_df_unique, by = "GENE") %>%
  filter(stability != "no IR")

#I dont want to have repeated rows (intros) for the same genes:
combined_exclusive_unique <- combined_exclusive %>%
  distinct(GENE, .keep_all = TRUE)
sum(duplicated(combined_exclusive_unique$GENE))

##here if I want to look at specific gene groups(but keep in mind that now introns are randomly selected)
stable4h_genes_only <- combined_exclusive_unique %>%
  filter(GENE_stability == "stable_4h")

##check relevant columns
subset_combined_exclusive <- combined_exclusive %>%
  dplyr::select(
    GENE, COORD, LENGTH,
    Cyt_A_UT_1.x, Cyt_B_UT_1.x, Cyt_A_30_1.x, Cyt_B_30_1.x, 
    Cyt_A_2h_1.x, Cyt_B_2h_1.x, Cyt_A_4h_1.x, Cyt_B_4h_1.x,
    Nuc_A_UT_1.x, Nuc_B_UT_1.x, Nuc_A_30_1.x, Nuc_B_30_1.x,
    Nuc_A_2h_1.x, Nuc_B_2h_1.x, Nuc_A_4h_1.x, Nuc_B_4h_1.x,
    Cyt_UT_mean, Cyt_30_mean, Cyt_2h_mean, Cyt_4h_mean,
    Nuc_UT_mean, Nuc_UT_TPM_mean, Nuc_30_TPM_mean, Nuc_2h_TPM_mean, Nuc_4h_TPM_mean,
    Cyt_UT_TPM_mean, Cyt_30_TPM_mean, Cyt_2h_TPM_mean, Cyt_4h_TPM_mean,
    Nuc_UT_IR_stability, Nuc_30_IR_stability, Nuc_2h_IR_stability, Nuc_4h_IR_stability,
    Cyt_UT_IR_stability, Cyt_30_IR_stability, Cyt_2h_IR_stability, Cyt_4h_IR_stability,
    stability, GENE_stability)

# --- STEP 2: Pivot the Data from Wide to Long Format ---
long_exclusive_data <- combined_exclusive_unique %>%
  pivot_longer(
    cols = c("Nuc_UT_TPM_stability", "Nuc_30_TPM_stability", 
             "Nuc_2h_TPM_stability", "Nuc_4h_TPM_stability",
             "Cyt_UT_TPM_stability", "Cyt_30_TPM_stability", 
             "Cyt_2h_TPM_stability", "Cyt_4h_TPM_stability"),
    names_to = c("Compartment", "Time"),
    names_pattern = "^(Nuc|Cyt)_([^_]+)_TPM_stability$",
    values_to = "TPM_stability") %>%
  mutate(Time = factor(Time, levels = c("UT", "30", "2h", "4h")))

# --- STEP 3: Remove the "no IR" Category and Set Factor Levels for 'stability' ---
# First, filter out rows where 'stability' is "no IR"
long_exclusive_data <- long_exclusive_data %>%
  filter(stability != "no IR") %>%  # Remove rows with "no IR"
  mutate(stability = factor(as.character(stability),
                            levels = c("unstable30min", "unstable2h", "unstable 4h", "stable 4h")))

# --- (Optional) Verify the Exclusive Gene Filtering ---
unique_genes_per_group <- long_exclusive_data %>%
  group_by(stability) %>%
  summarize(n_genes = n_distinct(GENE))
print(unique_genes_per_group)

# --- STEP 4: Plot the Mean TPM Stability Over Time, Color-Coded by Category ---
myStabColors <- colorRampPalette(c("darkgray", "darkred"))(4)

l <- ggplot(long_exclusive_data, aes(x = Time, y = TPM_stability, color = GENE_stability, group = GENE_stability)) +
  stat_summary(fun.data = mean_se, geom = "ribbon", aes(fill = GENE_stability), alpha = 0.2, color = NA) +  # Add ribbon first
  stat_summary(fun = mean, geom = "line", size = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  facet_wrap(~ Compartment, scales = "free_y") +
  labs(title = "RNA Stability Over Time (Exclusive Gene Groups)",
       x = "Time Point",
       y = "Mean TPM Stability (Relative to UT)") +
  coord_cartesian(ylim = c(0.0, 1.05)) +
  scale_color_manual(
    breaks = c("unstable_30min", "unstable_2h", "unstable_4h", "stable_4h"),
    values = myStabColors) +
  scale_fill_manual(  # Add this to match ribbon fill colors
    breaks = c("unstable_30min", "unstable_2h", "unstable_4h", "stable_4h"),
    values = myStabColors) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16))

l
###stability summary
# Summary: Mean TPM stability per group, time point, and compartment
# ---- Step 1: Summarize TPM stability ----
stability_summary <- long_exclusive_data %>%
  group_by(GENE_stability, Time, Compartment) %>%
  summarise(
    mean_TPM_stability   = mean(TPM_stability, na.rm = TRUE),
    median_TPM_stability = median(TPM_stability, na.rm = TRUE),
    sd_TPM_stability     = sd(TPM_stability, na.rm = TRUE),
    n                    = n(),
    .groups = "drop"
  )

# Separate for nucleus and cytoplasm
stability_summary_nuc <- stability_summary %>% filter(Compartment == "Nuc")
stability_summary_cyt <- stability_summary %>% filter(Compartment == "Cyt")

# ---- Step 2: Pairwise Wilcoxon tests (within each time point) ----
# Filter only nuclear data
data_for_stats <- long_exclusive_data %>%
  filter(Compartment == "Nuc")

# Perform pairwise comparisons at each time point
pairwise_pvals <- data_for_stats %>%
  group_by(Time) %>%
  nest() %>%
  ungroup() %>%
  mutate(
    pairwise_tests = map2(data, Time, ~ {
      pairwise_results <- combn(unique(.x$GENE_stability), 2, simplify = FALSE)
      map_dfr(pairwise_results, function(grps) {
        test_data <- .x %>% filter(GENE_stability %in% grps)
        w_test <- wilcox.test(TPM_stability ~ GENE_stability, data = test_data, exact = FALSE)
        tibble(
          Time = .y,
          group1 = grps[1],
          group2 = grps[2],
          p_value = w_test$p.value,
          statistic = w_test$statistic,
          method = w_test$method
        )
      })
    })
  ) %>%
  dplyr::select(pairwise_tests) %>%
  unnest(pairwise_tests)

# --- Step 1: Calculate the log2 Fold Change for Nuclear Enrichment ---
# First, filter out rows with "no IR", recalc log2FC, and set the factor levels for stability
combined_exclusive_filtered <- combined_exclusive %>%
  #filter(stability != "no IR") %>%
  mutate(
    nuc_enrichment_log2FC = log2(Nuc_UT_TPM_mean / Cyt_UT_TPM_mean),
    stability = factor(as.character(stability),
                       levels = c("unstable30min", "unstable2h", "unstable 4h", "stable 4h"))
  )

# Create the Violin Plot using a color grade (myStabColors) and allow negative values
p3 <- ggplot(combined_exclusive_filtered, aes(x = stability, y = nuc_enrichment_log2FC, fill = stability)) +
  geom_violin(trim = FALSE, alpha = 0.7) +                
  geom_boxplot(width = 0.1, outlier.shape = NA, fill = "white") +
  labs(
    title = "Violin Plot of Nuclear Enrichment log2FC",
    x = "Stability Group",
    y = "log2FC (Nuc_UT_TPM_mean / Cyt_UT_TPM_mean)"
  ) +
  scale_y_continuous(limits = c(-2.5, NA), expand = c(0, 0.05)) +
  scale_fill_manual(values = myStabColors) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16)
  )

print(p3)

# Summary statistics for log2FC by stability group
library(dplyr); library(tidyr); library(purrr); library(broom); library(readr)

# ---- Step 1: Compute summary statistics ----
log2fc_summary <- combined_exclusive_filtered %>%
  group_by(stability) %>%
  summarise(
    mean_log2FC   = mean(nuc_enrichment_log2FC, na.rm = TRUE),
    median_log2FC = median(nuc_enrichment_log2FC, na.rm = TRUE),
    sd_log2FC     = sd(nuc_enrichment_log2FC, na.rm = TRUE),
    min_log2FC    = min(nuc_enrichment_log2FC, na.rm = TRUE),
    max_log2FC    = max(nuc_enrichment_log2FC, na.rm = TRUE),
    n             = n(),
    .groups = "drop"
  )

# ---- Step 2: Pairwise Wilcoxon tests ----
# Get all pairwise combinations of stability groups
pairwise_groups <- combn(unique(combined_exclusive_filtered$stability), 2, simplify = FALSE)

# Run Wilcoxon tests for each pair
log2fc_pairwise <- map_dfr(pairwise_groups, function(grps) {
  test_data <- combined_exclusive_filtered %>% filter(stability %in% grps)
  w_test <- wilcox.test(nuc_enrichment_log2FC ~ stability, data = test_data, exact = FALSE)
  tibble(
    group1 = grps[1],
    group2 = grps[2],
    p_value = w_test$p.value,
    statistic = w_test$statistic,
    method = w_test$method
  )
})

```

##line plot for 50%
```{r}
# Step 1: Filter for 50% cutoff and reasonable TPMs
combined_dataTPMno100_50cut <- combined_data_50cut %>%
  filter(Nuc_UT_TPM_mean >= 7 & Cyt_UT_TPM_mean >= 7) %>%
  filter(is.na(Cyt_A_UT_1.x) | is.na(Cyt_B_UT_1.x) |
         (Cyt_A_UT_1.x < 99 & Cyt_B_UT_1.x < 99))

# Step 2: Merge with exclusive gene assignments (from 50% Venn analysis)
combined_exclusive_50cut <- combined_dataTPMno100_50cut %>%
  filter(GENE %in% gene_stability_df_unique_50cut$GENE) %>%
  left_join(gene_stability_df_unique_50cut, by = "GENE") %>%
  filter(stability != "no IR")

# Step 3: Keep only one intron per gene
combined_exclusive_unique_50cut <- combined_exclusive_50cut %>%
  distinct(GENE, .keep_all = TRUE)

# Subset relevant columns
subset_combined_exclusive_50cut <- combined_exclusive_50cut %>%
  dplyr::select(
    GENE, COORD, LENGTH,
    Cyt_A_UT_1.x, Cyt_B_UT_1.x, Cyt_A_30_1.x, Cyt_B_30_1.x, 
    Cyt_A_2h_1.x, Cyt_B_2h_1.x, Cyt_A_4h_1.x, Cyt_B_4h_1.x,
    Nuc_A_UT_1.x, Nuc_B_UT_1.x, Nuc_A_30_1.x, Nuc_B_30_1.x,
    Nuc_A_2h_1.x, Nuc_B_2h_1.x, Nuc_A_4h_1.x, Nuc_B_4h_1.x,
    Cyt_UT_mean, Cyt_30_mean, Cyt_2h_mean, Cyt_4h_mean,
    Nuc_UT_mean, Nuc_UT_TPM_mean, Nuc_30_TPM_mean, Nuc_2h_TPM_mean, Nuc_4h_TPM_mean,
    Cyt_UT_TPM_mean, Cyt_30_TPM_mean, Cyt_2h_TPM_mean, Cyt_4h_TPM_mean,
    Nuc_UT_IR_stability, Nuc_30_IR_stability, Nuc_2h_IR_stability, Nuc_4h_IR_stability,
    Cyt_UT_IR_stability, Cyt_30_IR_stability, Cyt_2h_IR_stability, Cyt_4h_IR_stability,
    stability, GENE_stability)

# Step 4: Pivot to long format for plotting
long_exclusive_data_50cut <- combined_exclusive_unique_50cut %>%
  pivot_longer(
    cols = c("Nuc_UT_TPM_stability", "Nuc_30_TPM_stability", 
             "Nuc_2h_TPM_stability", "Nuc_4h_TPM_stability",
             "Cyt_UT_TPM_stability", "Cyt_30_TPM_stability", 
             "Cyt_2h_TPM_stability", "Cyt_4h_TPM_stability"),
    names_to = c("Compartment", "Time"),
    names_pattern = "^(Nuc|Cyt)_([^_]+)_TPM_stability$",
    values_to = "TPM_stability") %>%
  mutate(Time = factor(Time, levels = c("UT", "30", "2h", "4h")),
         stability = factor(as.character(stability),
                            levels = c("unstable30min", "unstable2h", "unstable 4h", "stable 4h")))
long_exclusive_data_50cut$GENE_stability <- factor(
  long_exclusive_data_50cut$GENE_stability,
  levels = c("unstable_30min", "unstable_2h", "unstable_4h", "stable_4h")
)
# Step 5: Check gene distribution
long_exclusive_data_50cut %>%
  group_by(stability) %>%
  summarize(n_genes = n_distinct(GENE)) %>%
  print()

# Step 6: TPM stability over time plot
myStabColors <- colorRampPalette(c("darkgray", "darkred"))(4)
l_50cut <- ggplot(long_exclusive_data_50cut, aes(x = Time, y = TPM_stability, color = GENE_stability, group = GENE_stability)) +
  stat_summary(fun.data = mean_se, geom = "ribbon", aes(fill = GENE_stability), alpha = 0.2, color = NA) +
  stat_summary(fun = mean, geom = "line", size = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  facet_wrap(~ Compartment, scales = "free_y") +
  labs(title = "RNA Stability Over Time (Exclusive Gene Groups, 50% cutoff)",
       x = "Time Point", y = "Mean TPM Stability (Relative to UT)") +
  coord_cartesian(ylim = c(0.0, 1.05)) +
  scale_color_manual(values = myStabColors) +
  scale_fill_manual(values = myStabColors) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16)
  )

# Display the updated plot
print(l_50cut)

# ---- Step 1: Summarize TPM stability ----
stability_summary_50cut <- long_exclusive_data_50cut %>%
  group_by(GENE_stability, Time, Compartment) %>%
  summarise(
    mean_TPM_stability   = mean(TPM_stability, na.rm = TRUE),
    median_TPM_stability = median(TPM_stability, na.rm = TRUE),
    sd_TPM_stability     = sd(TPM_stability, na.rm = TRUE),
    n                    = n(),
    .groups = "drop"
  )

# Separate for nucleus and cytoplasm
stability_summary_nuc_50cut <- stability_summary_50cut %>% filter(Compartment == "Nuc")
stability_summary_cyt_50cut <- stability_summary_50cut %>% filter(Compartment == "Cyt")

# ---- Step 2: Pairwise Wilcoxon tests (within each time point) ----
# Filter only nuclear data
data_for_stats_50cut <- long_exclusive_data_50cut %>%
  filter(Compartment == "Nuc")

# Perform pairwise comparisons at each time point
pairwise_pvals_50cut <- data_for_stats_50cut %>%
  group_by(Time) %>%
  nest() %>%
  ungroup() %>%
  mutate(
    pairwise_tests = map2(data, Time, ~ {
      pairwise_results <- combn(unique(.x$GENE_stability), 2, simplify = FALSE)
      map_dfr(pairwise_results, function(grps) {
        test_data <- .x %>% filter(GENE_stability %in% grps)
        w_test <- wilcox.test(TPM_stability ~ GENE_stability, data = test_data, exact = FALSE)
        tibble(
          Time = .y,
          group1 = grps[1],
          group2 = grps[2],
          p_value = w_test$p.value,
          statistic = w_test$statistic,
          method = w_test$method
        )
      })
    })
  ) %>%
  dplyr::select(pairwise_tests) %>%
  unnest(pairwise_tests)

# Step 8: Violin plot of log2 nuclear enrichment
##use these just for GO and nuc enrichment (because these are the same ones that we had for lenght and GC
combined_dataTPMno100_50cut <- combined_data_50cut %>%
  filter(Nuc_UT_TPM_mean >= 5 & Cyt_UT_TPM_mean >= 1) %>%
  filter(is.na(Cyt_A_UT_1.x) | is.na(Cyt_B_UT_1.x) | 
  (Cyt_A_UT_1.x < 99 & Cyt_B_UT_1.x < 99))

# Step 2: Merge with exclusive gene assignments (from 50% Venn analysis)
combined_exclusive_50cut <- combined_dataTPMno100_50cut %>%
  filter(GENE %in% gene_stability_df_unique_50cut$GENE) %>%
  left_join(gene_stability_df_unique_50cut, by = "GENE") %>%
  filter(stability != "no IR")

combined_exclusive_filtered_50cut <- combined_exclusive_50cut %>%
  filter(stability != "no IR") %>%
  mutate(nuc_enrichment_log2FC = log2(Nuc_UT_TPM_mean / Cyt_UT_TPM_mean))

combined_exclusive_filtered_50cut <- combined_exclusive_filtered_50cut %>%
  mutate(stability = factor(stability, levels = c("unstable30min", "unstable2h", "unstable 4h", "stable 4h")))

p3_50cut <- ggplot(combined_exclusive_filtered_50cut, aes(x = stability, y = nuc_enrichment_log2FC, fill = stability)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.1, outlier.shape = NA, fill = "white") +
  labs(title = "Violin Plot of Nuclear Enrichment log2FC (50% cutoff)",
       x = "Stability Group", y = "log2FC (Nuc_UT_TPM / Cyt_UT_TPM)") +
  scale_y_continuous(limits = c(-2.5, NA), expand = c(0, 0.05)) +
  scale_fill_manual(values = myStabColors) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line.x = element_line(color = "black"),
        axis.line.y = element_line(color = "black"),
        axis.ticks = element_line(color = "black"),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 16))

print(p3_50cut)

```
